**Machine learning identifies unique taxa differentiating proximal and distal

### 2017-05-01 submission/manuscript.Rmd
manuscript and oncogenic otu code added

#### Microbial membership and diversity of the proximal and distal colon 
Proximal and distal lumenal and mucosal samples were collected from 20 healthy humans that had not undergone bowel preparation prior to the procedure (Figure 1). In addition, individuals collected stool at home one week prior to the procedure. To characterize the bacterial communities present at these sites, 16S rRNA gene sequencing was performed on extracted DNA from each sample. The relative abundance of each phyla for each site is shown in Figure 2A. This data combines the samples from all participants in the study. Each site is primarily dominated by Firmicutes and Bacteriodetes, consistent with known variability in human microbiome research. ((Is anything statistically significant here? use Doherty's code)). Likewise, samples had varying levels of diversity at each site, irrespective of the individual (Figure 2B). That is, while some individuals had a more diverse right mucosa, some had a more diverse left mucosa, and so on. Thus there is no clear pattern of changes in microbial diversity along the gut axis.

@@

To compare similarity between sides or sites, we calculated theta YC distances from OTU abundances and compared distances between sies for all patients. Again, across all patient samples we saw a range of theta YC distances when comparing sample locations (Figure 3A) and again those ranges did not follow a clear pattern on an individual basis. However, when comparing median distances between the right lumen and mucosa, the left versus right lumen, the left versus right mucosa and the left lumen and mucosa, we found that the right lumen and mucosa were most similar to each other than the other samples (P < 0.005, Wilcoxon, BH adjustment). Next, we calculated thetaYC distances to examine how each sample compares to the stool at exit. Amidst variability between patients, we did identify significantly smaller thetaYc distance between the left lumenal sample and the exit stool (P < 0.05, Wilcoxon, BH adjustment). Further, there was an even more divergent difference in the comparisons of the left mucosa to the exit stool, indicating that the mucosa is quite different from the stool as compared to lumen (P < 0.0005, Wilcoxon, BH adjustment). This indicates that the contents of the left lumen are most representative of stool at exit, and the microbes remaining on the mucosa are much different. ((mention the other 2 comparisons? "among the four comparisons made..."))

#### Random Forest classification models identify important OTUs on each side
To identify OTUs that are distinct at each biogeographical site, we constructed several Random Forest machine-learning models built on OTU abundances. The first model we built to classify the lumen versus mucosal samples for the right and left sides, independently (Figure 4A). The constructed model used ((Xopt? need to follow up with this too)) features for the left and ((Xopt)) for the right. The models performed very well when classifying these samples (AUC X and Y). The OTUs that were most predictive of each site are identified by their greatest mean decrease in accuracy when removed from the model. For distinguishing the right lumen and mucosa, OTUs from the _Bacteriodes_, _Actinomyces_, _Psuedomonas_ and two OTUs from the _Enterobacteraceae_ genera were differentially abundant (Figure 4B). The model classifying the left lumen and mucosa identified OTUs from _Turicibacter_, _Anaerococcus_, _Finegoldia_, _Peptoniphilus_ and two OTUs from the _Anaerococcus_ genera that could distinguish lumen from mucosal samples (Figure 4B). These results indicate that there are fine differences between the different sites of the colon, and that these can be traced down to specific OTUs on each side.

Next, we built a model to differentiate the left and right microbiome. The model performed best when distinguishing the left versus right mucosa (Figure 5A) compared to the left versus right lumen. The model was able to explain ((X%)) of the variance. OTUs that were differentially abundant between the left and right mucosa included members of the _Porphyromonas_, _Murdochiella_, _Finegoldia_, _Anaerococcus_ and _Peptoniphilus_ genera (Figure 5B). The model was less-effective at classifying the right and left lumen ((AUC)). However it did identify differentially abundant OTUs such as three members of the _Bacteroides_ genus, a _Clostridium IV_ OTU and an _Oscillibacter_ OTU (Figure 5C).

- There is some commonality between the OTUs - drivers of left sidedness are consistent in both lumen and mucosa (more of this in discussion)

@@

### Discussion

@@

-how did the analyses answer the central question?
The analyses identified OTUs that are relatively unique to each location, among a wide variation differences between study participants

>>>>

#### Microbial membership and diversity of the proximal and distal colon 
Proximal and distal lumenal and mucosal samples were collected from 20 healthy humans that had not undergone bowel preparation prior to the procedure (Figure 1). Participants also collected stool at home one week prior to the procedure. To characterize the bacterial communities present at these sites, 16S rRNA gene sequencing was performed on extracted DNA from each sample. The relative abundance of each phyla for each site is shown in Figure 2A. This data combines the samples from all participants in the study. Each site is primarily dominated by Firmicutes and Bacteriodetes, consistent with known variability in human microbiome research. ((Is anything statistically significant here? use Doherty's code)). Likewise, samples had varying levels of diversity at each site, irrespective of the individual (Figure 2B). That is, while some individuals had a more diverse right mucosa, some had a more diverse left mucosa, and so on. Thus there was no clear pattern of changes in microbial diversity along the gut axis.

@@

To compare similarity between sides (left or right) or sites (lumen or mucosa), we calculated theta YC distances from OTU abundances and compared distances for all individuals. Again, across all patient samples we saw a range of theta YC distances when comparing sample locations (Figure 3A) and again those ranges did not follow a clear pattern on an individual basis. However, when comparing median distances between the right lumen and mucosa, the left versus right lumen, the left versus right mucosa and the left lumen and mucosa, we found that the right lumen and mucosa were most similar to each other than the other samples (P < 0.005, Wilcoxon, BH adjustment). Next, we calculated thetaYC distances to examine how each sample compared to the home-collected exit stool. Amidst variability between patients, we did identify significantly smaller thetaYc distance between the left lumenal sample and the exit stool (Figure 3B, P < 0.05, Wilcoxon, BH adjustment). Further, there was an even larger difference in the comparisons of the left mucosa to the exit stool, indicating that the mucosa is quite different from the stool as compared to lumen (P < 0.0005, Wilcoxon, BH adjustment). This finding suggests that the contents of the left lumen are most representative of stool at exit, and the microbes remaining on the mucosa are much different. ((mention the other 2 comparisons? "among the four comparisons made..."))

#### Random Forest classification models identify important OTUs on each side
To identify OTUs that are distinct at each biogeographical site, we constructed several Random Forest machine-learning models built on OTU abundances. The first model we built to classify the lumen versus mucosal samples for the right and left sides, independently (Figure 4A). The constructed model used ((Xopt? need to follow up with this too)) features for the left and ((Xopt)) for the right. The models performed very well when classifying these samples (AUC X and Y). The OTUs that were most predictive of each site are identified by their greatest mean decrease in accuracy when removed from the model. For distinguishing the right lumen and mucosa, OTUs from the _Bacteriodes_, _Actinomyces_, _Psuedomonas_ and two OTUs from the _Enterobacteraceae_ genera were differentially abundant (Figure 4B). The model classifying the left lumen and mucosa identified OTUs from _Turicibacter_, _Finegoldia_, _Peptoniphilus_ and two OTUs from the _Anaerococcus_ genera that could distinguish lumen from mucosal samples (Figure 4B). These results indicate that there are fine differences between the different sites of the colon, and that these can be traced down to specific OTUs on each side.

Next, we built a model to differentiate the left and right microbiome. The model performed best when distinguishing the left versus right mucosa (Figure 5A) compared to the left versus right lumen. The model was able to explain ((X%)) of the variance. OTUs that were differentially abundant between the left and right mucosa included members of the _Porphyromonas_, _Murdochiella_, _Finegoldia_, _Anaerococcus_ and _Peptoniphilus_ genera (Figure 5B). The model was less-effective at classifying the right and left lumen ((AUC)). However it did identify differentially abundant OTUs such as three members of the _Bacteroides_ genus, a _Clostridium IV_ OTU and an _Oscillibacter_ OTU (Figure 5C). This analysis found that some of the same OTUs that are distinct between the mucosa and lumen also drive sidedness- such as _Anaerococcus_ and _Finegoldia_.

@@

### Discussion
Here we identified bacterial taxa that are specific to the lumen and mucosa of the right and left human colon from samples collected during unprepared colonoscopy. We found that all locations contained a range of phyla and a range of diversity, but that there was a wide variability between subjects. Pairwise comparisons of each of the sites revealed that the right mucosa and lumen were most similar to each other. Further, comparison of colonoscopy-collected samples with samples collected from stool at home showed that the left lumen is most similar to stool at exit. Random Forest algorithms built on OTU abundances from each sample identified microbes that are particular to each location of the colon.

@@

Put this stuff into context with Sears PNAS paper

Fuso found on the right side?
